Phylogenetics and the human microbiome.

### 2013-09-24 sbreview.tex
Traits and evolution.

Why, as systematic biologists, should we be interested in what these microbial ecologists have done?
System with a lot of data and a lot of questions.
Indeed, just to do ecology you have to do sequencing.
Many of the issues that are raised by high throughput sequencing of beasties that are more commonly of interest for phylogenetics folk are present, and worse, in microbial studies.
Taxonomic issues.

Lack of "obvious" ways of grouping beasties has led to innovation.
It means something that they have developed their own way of using trees, although there is some overlap with previous literature.
Questions of microbial genomics, translated into questions of function, have real consequences.

@@

Opportunities for SB audience to contribute to this field.

@@

As we will see, 16S is used frequently as a way of measuring diversity.
Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees (cite PI-CRUST).

@@

Metagenomic data is typically used to infer information about metabolic capacity (cite Humann) rather than make statements of an evolutionary nature.
One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
This is the approach taken by the teams of Christian von Mering \cite{von2007quantitative,stark2010mltreemap} and Jonathan Eisen \cite{wu2008amphora} (in preparation).
Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.

@@

\section{Phylogenetic challenges for human microbiome research}

@@

@@

@@

But in addition to deeper probing we can also expect wider coverage.
Lots of data.
Citizen microbiome

@@

@@

>>>>

Why, as systematic biologists, should we be interested in what these microbial ecologists have done?
System with a lot of data and a lot of questions.
Microbes can only be identified by their molecular sequence.
Indeed, just to do ecology you have to do sequencing.
Many of the issues that are raised by high throughput sequencing of beasties that are more commonly of interest for phylogenetics folk are present, and worse, in microbial studies.

Taxonomic issues.
Lack of "obvious" ways of grouping beasties.
Lack of broad phenotypic characterization.
These issues motivate sequence- and evolution-based methodology.

This has led to innovation in ways of using trees.
It means something that they have developed their own way of using trees, although there is some overlap with previous literature.
Questions of microbial genomics, translated into questions of function, have important consequences for human health.

@@

In this review I will describe human microbiome research from the perspective of a phylogenetics researcher.
I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees and the challenges that human microbiome research poses for the phylogenetics community.
I will finish with opportunities for SB audience to contribute to this field.

@@

As we will see, 16S is used frequently as a way of measuring diversity.
Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} using either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.

@@

Metagenomic data is typically used to infer information about metabolic capacity (cite Humann) rather than make statements of an evolutionary nature.
One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
This is the approach taken by \citep{von2007quantitative,stark2010mltreemap} and Jonathan Eisen \citep{wu2008amphora} (in preparation).
Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.

@@

\section{Phylogenetic challenges and opportunities in human microbiome research}

@@

There is a strange divide between trait evolution methods for macro- and micro-fauna.
The results of simple methods are reasonable.
Possible that improved methods, perhaps involving whole-genome evolutionary modeling, could shed light on the problem.

@@

In any case, what about trait evolution models that take HGT into account?
What about trait evolution models that are on KEGG?

@@

But in addition to deeper probing we can also expect wider coverage.
Lots of data.
Citizen microbiome projects.

@@

Perhaps there is a role for trait evolution methods, or perhaps a simple method does prediction to within how good the data is.
There is the possibility that we don't have to think about individual organisms at all, but rather can think about the collection of genes and their metabolic network as a meta-organism (cite Elbo).
Question: to what extent is modeling cellular-level evolutionary history interesting?

@@

\section{Acknowledgements}
Aaron Darling, Connor McCoy,
FHCRC folk.
ATD grant.
